\Title

### 2013-03-01 synmut.tex
supplementary figures correct numbering
Little changes here and there

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious.  We
argue that synonymous mutations reach high frequencies via genetic hitchhiking
due to limited recombination in HIV-1 populations~\citep{neher_recombination_2010,
batorsky_estimate_2011}. We then compare our observations to computational
models of HIV-1 evolution and derive estimates for the effect of synonymous mutations
 on fitness.  Extending the analysis of fixation probabilities to the
nonsynonymous mutations, we show that time dependent selection or strong
competition of escape mutations inside the same epitope are necessary to explain
the observed patterns of fixation and loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency. A neutral mutation
segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e., with
probability $1-\nu$, it goes extinct. As illustrated in the inset of
\FIG{aftsyn}, this is a  consequence of the fact that (i) exactly one
individual in the current population will be the common ancestor of the entire
future population at a particular site and (ii) this ancestor has a probability
$\nu$ of carrying the mutation (assuming the neutral mutation is not
preferentially associated with genomes of high or low fitness). Deleterious or
beneficial mutations fix less or more often than neutral ones, respectively.
\FIG{aft} shows the time course the frequencies of all synonymous and
nonsynonymous mutations observed \env, \shankaregion, in patient
p10~\citep{shankarappa_consistent_1999}, respectively. Despite many synonymous
mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
Strictly speaking, no mutation in the HIV-1 population ever fixes because the 
mutation rate and the population size are large. We define ``fixation'' or ``loss''
by not observing the mutation in the sample.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
mutation is simply the likelihood that the future common ancestor at this 
position is currently carrying it, i.e., the mutation frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in \shankaregion{} tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for lack of fixation of synonymous mutations in
\shankaregion are secondary structures in the viral RNA, the disruption of which
is deleterious to the virus \citep{forsdyke_reciprocal_1995,
snoeck_mapping_2011, sanjuan_interplay_2011}.

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV-1 genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems.  We aligned the within-patient sequence samples 
to the reference NL4-3 strain used in ref.~\citep{watts_architecture_2009} and 
thereby assigned SHAPE reactivities to most positions in the alignment. 
We then calculated the distributions of SHAPE reactivities for synonymous 
polymorphisms that fixed or where subsequently lost (only polymorphisms with 
frequencies above 15\%).
As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
histogram) are systematically larger than of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population. The average over all
mutations that are not observed (green) lies between  those that fix and
those that get lost. Note that this analysis will be sensitive only at positions
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

@@

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome. This region, \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious mutations
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
\FIG{aft}, which shows that many mutations change rapidly in frequency as a
flock.

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, which shows the distribution of times after which synonymous
alleles at intermediate frequencies fixed or were lost. The
typical time to loss is of the order of 500 days. If this loss is driven by the
deleterious effect of the mutation, this corresponds to deleterious effects $s_d$ of
the order of $- 0.002$ per day.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of  synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
day). Only simulations that reproduce the synonymous diversity and fixation
patterns observed in data are shown. These simulations demonstrate that
deleterious effects are around $-0.002$ and a large fraction of the 
synonymous mutations needs to be deleterious. As expected, larger
$s_d$ require larger $\epsilon$. Parameters are chosen
from prior distributions uniform in logspace as indicated by the red box (see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

@@

\FIG{simfixpvar} shows simulations results for the fixation probability and the
synonymous diversity for different deleterious effects of synonymous mutations.
We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
with an allele at frequency $0.25 < \nu < 0.75$. The synonymous diversity
observed in patient data is indicated in the figure.  To quantify the depression
of the fixation probability, we calculate the area between the measured fixation
probability and the diagonal, which is the neutral expectation
(\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
$-0.5$; if every mutation fixes, the area will be $+0.5$. In HIV-1 infected
patients, we find $A_\text{syn} \approx -0.2$ for synonymous changes and
$A_\text{nonsyn} \approx 0$ for nonsynonymous changes. In the three simulations
shown in \FIG{simfixpvar}, the fixation probability of synonymous alleles
decreases from the neutral expectation ($A_\text{syn} \approx 0$) to zero
($A_\text{syn} \approx -0.5$) as their fitness cost increases; the synonymous
diversity plummets as well, as deleterious mutations are selected against.

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the parameters in
certain bounds, see \FIG{simsfig}. Among all simulations, we select the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous mutations but,
simultaneously, a moderately high synonymous diversity. Specifically,
\FIG{simsfig} shows parameter combinations for which we found $A_\text{syn} <
-0.15$ and $0.0025 < P_\text{interm} < 0.010$. These conditions indicate that a
high fraction ($\gtrsim 0.8$) of sites has to be deleterious with effect size
$|s_d| \sim 0.002$.  This result fits well the expectation based on the
fixation/extinction times above (see \FIG{fixp1}). The results are plausible:
(i) a substantial depression in $\pfix$ requires pervasive deleterious
mutations, otherwise the majority of observed polymorphisms are neutral and no
depression is observed; (ii) in order to hitchhike, the deleterious effect size
has to be much smaller than the escape rate, otherwise the double mutant has
little or no fitness advantage. Consistent with this argument, larger
deleterious effects in \FIG{simsfig} correspond to larger escapes rates. (iii)
mutations with a deleterious effect smaller than approximately $0.001$ behave
neutrally consistent with the typical coalescent times observed in HIV-1.

The above simulation show that hitchhiking can explain the observation of
deleterious mutations that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, they almost
always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue line in \FIG{fixp}: in an HIV-1
infection, nonsynonymous mutations at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous mutations suggests the rapid rise and fall of many alleles. We
test two possible mechanisms that are biologically plausible and could explain
the transient rise of nonsynonymous mutations: time-dependent selection and
within-epitope competition.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of mutations in longitudinal data of HIV-1 \env{} evolution,
we demonstrate selection against synonymous substitutions in the comparatively
conserved regions C2-C5 of the \env{} gene. Comparison with biochemical studies
of base pairing propensity in RNA genome of HIV-1 indicates that these
mutations are deleterious, at least in part, because they disrupt stems in RNA
secondary structures. Computational modeling shows that these mutations have
deleterious effects on the order of $0.002$ and that they are brought to high
frequency through linkage to adaptive mutations.

The fixation and extinction times and probabilities represent a rich and simple
summary statistics useful to characterize longitudinal sequence data and compare
to models via computer simulations. A method that is similar to ours {\it in
spiritu} has been recently used in a longitudinal study of influenza
evolution~\citep{strelkowa_clonal_2012}. The central quantity used in this
article, however, is a ratio between propagators of nonsynonymous and synonymous
mutations. The latter is used as an approximately neutral control; this method
can therefore not be used to investigate synonymous changes themselves. More
generally, evolutionary rates at synonymous sites are often used as a baseline
to detect purifying or diversifying selection at the protein level
\cite{Hurst:2002p32608}. It has been pointed out, however, that the rate of
evolution at synonymous sites varies considerable along the HIV-1 genome
\citep{mayrose_towards_2007} and that this variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}.

@@

While classical population genetics assumes that the dominant stochastic force
is genetic drift, i.e., non-heritable fluctuations in offspring number, our
results show that stochasticity due to linked selection is much more important.
Such fluctuations have been termed \emph{genetic draft} by
\citet{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
such as HIV-1 has been characterized in \citep{neher_genetic_2011}. Importantly,
large population sizes are compatible with low diversity and fast coalescence
when draft dominates over drift.

Contrary to na\"ive expectations, the adaptive escape mutations do not seem to
be unconditionally beneficial. Otherwise we would observe almost sure fixation
of nonsynonymous mutations once they reach intermediate frequencies. Instead, we
find that the fixation probability of nonsynonymous mutations is roughly given
by its frequency. There are several possible explanations for this observation.
Similar to synonymous mutations, the majority of nonsynonymous mutations could
be weakly deleterious and the adaptive and deleterious parts conspire to yield a
more neutral-like averaged fixation probability. While weakly deleterious 
nonsynonymous mutation certainly exist and will contribute to a depression of the
fixation probability, we have seen that a substantial depression requires that
weakly deleterious non-synonymous polymorphisms at high frequency greatly 
outnumber escape mutations. This seems unlikely, since non-synonymous diversity
exceeds synonymous diversity despite the overall much greater constraint on
the aminoacid sequence.

Alternatively, the lack of fixation could be due to time dependent environment
through an immune system that is catching up, or competition between mutations
that mediate escape within the same epitope. We explore both of these
possibilities and find that both produce the desired effect. Furthermore, there
is experimental evidence in support of both of these hypotheses. Serum from HIV-1
infected individuals typically neutralizes the virus that dominated the
population a few (3-6) month earlier \citep{richman_rapid_2003}. This suggests that
escape mutations cease to be beneficial after a few months and might revert if
they come with a fitness cost. Deep sequencing of regions of \env{} after
antibody escape have revealed multiple escape mutations in the same epitope
\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
mutations is sufficient for escape but most combinations of them do not provide
any additional benefit to the virus. Hence only one mutation will spread and the
others will be driven out of the population although they transiently reach high
frequencies. The rapid emergence of multiple escape mutations in the same
epitope implies a large effective population size that explores all necessary point
mutations rapidly. A similar point has been made recently by Boltz {\it et al.}
in the context of preexisting drug resistance mutations
\citep{boltz_ultrasensitive_2012}.

Our results emphasize the inadequacy of independent site models of HIV-1 evolution
and the common assumption that selection is time independent or additive. 
If genetic variation is only transiently beneficial, existing estimates of the
strength of selection \citep{neher_recombination_2010,batorsky_estimate_2011}
could be substantial underestimates. Furthermore, weak conservation and
time dependent selection results in estimates of evolutionary 
rates that depend on the time interval of observation with lower rates across
larger intervals. This could mean that deep nodes in phylogenies are older than 
they appear.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We would like to thank Jan Albert, Trevor Bedford and Pleuni Pennings for 
stimulating discussions and critical reading of the manuscript.
This work is supported by the ERC starting grant HIVEVO 260686 and 
in part by the National Science Foundation under Grant No.~NSF PHY11-25915.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{natbib}
\bibliography{bib}
\newpage
\appendix
\onecolumngrid
\include{supplement_body}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious.  We
argue that synonymous mutations reach high frequencies via genetic hitchhiking
due to limited recombination in HIV-1 populations~\citep{neher_recombination_2010,
batorsky_estimate_2011}. We then compare our observations to computational
models of HIV-1 evolution and derive estimates for the effect of synonymous mutations
 on fitness.  Extending the analysis of fixation probabilities to the
nonsynonymous mutations, we show that time-dependent selection or strong
competition of escape mutations inside the same epitope are necessary to explain
the observed patterns of fixation and loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency. A neutral mutation
segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. As illustrated in the inset of
\FIG{aftsyn}, this is a  consequence of the fact that (i) exactly one
individual in the current population will be the common ancestor of the entire
future population at a particular site and (ii) this ancestor has a probability
$\nu$ of carrying the mutation (assuming the neutral mutation is not
preferentially associated with genomes of high or low fitness). Deleterious or
beneficial mutations fix less or more often than neutral ones, respectively.
\FIG{aft} shows the time course the frequencies of all synonymous and
nonsynonymous mutations observed \env, \shankaregion, in patient
p10~\citep{shankarappa_consistent_1999}, respectively. Despite many synonymous
mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
Strictly speaking, no mutation in the HIV-1 population ever fixes because the 
mutation rate and the population size are large. We define ``fixation'' or ``loss''
by not observing the mutation in the sample.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
mutation is simply the likelihood that the future common ancestor at this 
position is currently carrying it, i.e. the mutation frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in \shankaregion{} tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for lack of fixation of synonymous mutations in
\shankaregion{} are secondary structures in the viral RNA, the disruption of which
is deleterious to the virus \citep{forsdyke_reciprocal_1995,
snoeck_mapping_2011, sanjuan_interplay_2011}.

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV-1 genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems.  We aligned the within-patient sequence samples 
to the reference NL4-3 strain used by \citet{watts_architecture_2009} and 
thereby assigned SHAPE reactivities to most positions in the alignment. 
We then calculated the distributions of SHAPE reactivities for synonymous 
polymorphisms that fixed or where subsequently lost (only polymorphisms with 
frequencies above 15\%).
As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
histogram) are systematically larger than of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population. The average over all
mutations that are not observed (green) lies between  those that fix and
those that get lost. Note that this analysis will be sensitive only at positions
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

@@

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}, compare the red
triangle line versus the green square line in \FIG{fixp2}).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious mutations
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
\FIG{aft}, which shows that many mutations change rapidly in frequency as a
flock.

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, which shows the distribution of times after which synonymous
alleles at intermediate frequencies become fixed or lost. The typical time to
loss is of the order of 500 days. If this loss is driven by the deleterious
effect of the mutation, this corresponds to deleterious effects $s_d$ of the
order of $- 0.002$ per day. (This is only an average estimate: every single
mutation is expected to have a slightly different fitness effect.)

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
day). Only simulations that reproduce the synonymous diversity and fixation
patterns observed in data are shown. These simulations demonstrate that
deleterious effects are around $-0.002$ and a large fraction of the 
synonymous mutations needs to be deleterious. As expected, larger
$s_d$ require larger $\epsilon$. Parameters are chosen
from prior distributions uniform in logspace as indicated by the red rectangle
(see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

@@

\FIG{simfixpvar} shows simulations results for the fixation probability and the
synonymous diversity for different deleterious effects of synonymous mutations.
We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
with an allele at frequency $0.25 < \nu < 0.75$. The synonymous diversity
observed in patient data is indicated in the figure.  To quantify the depression
of the fixation probability, we calculate the area between the measured fixation
probability and the diagonal, which is the neutral expectation
(\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
$-0.5$; if every mutation fixes, the area will be $+0.5$. In HIV-1 infected
patients, we find $P_\text{interm} \approx 0.005$, $A_\text{syn} \approx -0.2$
for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
changes. In the three simulations shown in \FIG{simfixpvar}, the fixation
probability of synonymous alleles decreases from the neutral expectation
($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.5$) as their
fitness cost increases; the synonymous diversity plummets as well, as
deleterious mutations are selected against.

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the parameters in
certain bounds, see \FIG{simsfig}. Among all simulations, we select the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous mutations but,
simultaneously, a moderately high synonymous diversity. Specifically,
\FIG{simsfig} shows parameter combinations for which we found $A_\text{syn} <
-0.15$ and $0.0025 < P_\text{interm} < 0.010$. These conditions indicate that a
high fraction ($\gtrsim 0.8$) of sites has to be deleterious with effect size
$|s_d| \sim 0.002$.  This result fits well the expectation based on the
fixation/extinction times above (see \FIG{fixp1}). The results are plausible:
(i) a substantial depression in $\pfix$ requires pervasive deleterious
mutations, otherwise the majority of observed polymorphisms are neutral and no
depression is observed; (ii) in order to hitchhike, the deleterious effect size
has to be much smaller than the escape rate, otherwise the double mutant has
little or no fitness advantage. Consistent with this argument, larger
deleterious effects in \FIG{simsfig} correspond to larger escapes rates; and (iii)
mutations with a deleterious effect smaller than approximately $0.001$ behave
neutrally, consistently with the typical coalescent times observed in HIV-1.

The above simulation show that hitchhiking can explain the observation of
deleterious mutations that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, they almost
always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
infection, nonsynonymous mutations at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous mutations suggests the rapid rise and fall of many alleles. We
test two possible mechanisms that are biologically plausible and could explain
the transient rise of nonsynonymous mutations: time-dependent selection and
within-epitope competition.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of mutations in longitudinal data of HIV-1 \env{} evolution,
we demonstrate selection against synonymous substitutions in the comparatively
conserved regions C2-C4 of the \env{} gene. Comparison with biochemical studies
of base pairing propensity in RNA genome of HIV-1 indicates that these
mutations are deleterious, at least in part, because they disrupt stems in RNA
secondary structures. Computational modeling shows that these mutations have
deleterious effects on the order of $0.002$ and that they are brought to high
frequency through linkage to adaptive mutations.

The fixation and extinction times and probabilities represent a rich and simple
summary statistics useful to characterize longitudinal sequence data and compare
to models via computer simulations. A method that is similar to ours {\it in
spiritu} has been recently used in a longitudinal study of influenza
evolution~\citep{strelkowa_clonal_2012}. The central quantity used in that
article, however, is a ratio between propagators of nonsynonymous and synonymous
mutations. The latter is used as an approximately neutral control; this method
can therefore not be used to investigate synonymous changes themselves. More
generally, evolutionary rates at synonymous sites are often used as a baseline
to detect purifying or diversifying selection at the protein level
\cite{Hurst:2002p32608}. It has been pointed out, however, that the rate of
evolution at synonymous sites varies considerable along the HIV-1 genome
\citep{mayrose_towards_2007} and that this variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}.

@@

While classical population genetics assumes that the dominant stochastic force
is genetic drift, i.e. non-heritable fluctuations in offspring number, our
results show that stochasticity due to linked selection is much more important.
Such fluctuations have been termed \emph{genetic draft} by
\citet{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
such as HIV-1 has been characterized in \citep{neher_genetic_2011}. Importantly,
large population sizes are compatible with low diversity and fast coalescence
when draft dominates over drift.

Contrary to na\"ive expectations, the adaptive escape mutations do not seem to
be unconditionally beneficial. Otherwise we would observe almost sure fixation
of nonsynonymous mutations once they reach intermediate frequencies. Instead, we
find that the fixation probability of nonsynonymous mutations is roughly given
by its frequency. There are several possible explanations for this observation.
Similar to synonymous mutations, the majority of nonsynonymous mutations could
be weakly deleterious and the adaptive and deleterious parts conspire to yield a
more neutral-like averaged fixation probability. While weakly deleterious 
nonsynonymous mutation certainly exist and will contribute to a depression of the
fixation probability, we have seen that a substantial depression requires that
weakly deleterious nonsynonymous polymorphisms at high frequency greatly 
outnumber escape mutations. This seems unlikely, since nonsynonymous diversity
exceeds synonymous diversity despite the overall much greater constraints on
the amino acid sequence.

Alternatively, the lack of fixation could be due to time-dependent environment
through an immune system that is catching up, or competition between mutations
that mediate escape within the same epitope. We explore both of these
possibilities and find that both produce the desired effect in computer models. Furthermore, there
is experimental evidence in support of both of these hypotheses. Serum from HIV-1
infected individuals typically neutralizes the virus that dominated the
population a few (3-6) month earlier \citep{richman_rapid_2003}. This suggests that
escape mutations cease to be beneficial after a few months and might revert if
they come with a fitness cost. Deep sequencing of regions of \env{} after
antibody escape have revealed multiple escape mutations in the same epitope
\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
mutations is sufficient for escape but most combinations of them do not provide
any additional benefit to the virus. Hence only one mutation will spread and the
others will be driven out of the population although they transiently reach high
frequencies. The rapid emergence of multiple escape mutations in the same
epitope implies a large effective population size that explores all necessary point
mutations rapidly. A similar point has been made recently by Boltz {\it et al.}
in the context of preexisting drug resistance mutations
\citep{boltz_ultrasensitive_2012}.

Our results emphasize the inadequacy of independent site models of HIV-1 evolution
and the common assumption that selection is time independent or additive. 
If genetic variation is only transiently beneficial, existing estimates of the
strength of selection \citep{neher_recombination_2010,batorsky_estimate_2011}
could be substantial underestimates. Furthermore, weak conservation and
time-dependent selection results in estimates of evolutionary 
rates that depend on the time interval of observation with lower rates across
larger intervals. This implies that deep nodes in phylogenies might be older than 
they appear.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We thank Jan Albert, Trevor Bedford and Pleuni Pennings for 
stimulating discussions and critical reading of the manuscript.
This work is supported by the ERC starting grant HIVEVO 260686 and 
in part by the National Science Foundation under Grant No.~NSF PHY11-25915.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{natbib}
\bibliography{bib}
\newpage
\appendix
\onecolumngrid
\setcounter{figure}{0}
\include{supplement_body}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
